---
title: "Main data processing script"
output: html_notebook
Authors: GLL, AF, EB, MW, AD, IE
---

 

```{r}

#Filtering BIOtime (collated pairs)
library(tidyverse) # a suite of tidy packages useful for data manipulation
library("dplyr") # data manipulation
library("tibble") # help to create simple dataframes
library("readr") # flexible package for reading in data files
library("ggplot2") # data visualization
library("magrittr") # aids in syntax and piping
library(progress)
library('rglobi')
library(taxize) # for taxon names

```


```{r}

#load data
load("data/prep_biotime/collated_pairs.RData") #full dataset with species/abundances
load('data/prep_biotime/bio_pairs_10km.RData') #"key" with overlapping studies, and reference to species in collated.pairs
load('data/prep_biotime/meta_pairs_10km.RData') #biotime metadata for 10 km pairs to use
#subseted<-collated.pairs[1:10000,]

#load functions
source("functions_data.processing.R")

```


Add taxa ID so that easy to ID which genera belong to what type of animal/plant/fungi:

```{r}

collated.pairs$ID <- as.character(collated.pairs$ID)
meta.pairs$STUDY_ID <- as.character(meta.pairs$STUDY_ID)
collated.pairs <- left_join(collated.pairs, meta.pairs[, c(1, 3,4, 12, 13)], by=c("ID"="STUDY_ID"))

head(collated.pairs)

```







# 1. Select species with >3 observations across studies and years

Remove species with less than three observation across all studies and years and save them in a list:

```{r}

low_sampling <- collated.pairs %>%
  group_by(GENUS_SPECIES) %>% #group by species
  summarize(n=n()) %>% #count the number of each species
  arrange(n)%>% #arrange in ascending order
  filter(n<3) #only get species with less than three observations
#across all studies and years 


low_sampled_species_list <- low_sampling$GENUS_SPECIES #get list of low sampled species
filtered.collated.pairs <- collated.pairs %>% #get original collated pairs dataset
  filter(!GENUS_SPECIES %in% low_sampled_species_list) #filter out observations with rare species

```








# 2. Standarize abundances by sampling effort


Visualize variation in sampling effort across years within each study:

Plot the number of plots per study (colors) across years:

```{r}

temp <- filtered.collated.pairs %>% 
  group_by(ID, YEAR) %>% #group by study, year
  summarize(EFFORT.YEAR = n_distinct(PLOT)) #count number of distinct plots

plot <- ggplot(data = temp, aes(x = YEAR, y = EFFORT.YEAR, group = ID, color = ID)) +
  theme_classic() +
  geom_line();plot

```


Zooming in on studies other than those 4 studies:

```{r}

plot2 <- temp %>% 
  filter(!(ID %in% c(54, 295, 296, 355))) %>% 
  ggplot(aes(x = YEAR, y = EFFORT.YEAR, group = ID, color = ID)) + 
  theme_classic() +
  geom_line();plot2

```

We can se lots of variation in sampling effort (plot number) among years here too




### 2.1. Standarize abundance and biomass

Back at the species level, standardize abundances by sum.allrawdata.ABUNDANCE or BIOMASS / EFFORT.YEAR:

```{r}

collated.pairs_standardized <- filtered.collated.pairs %>% 
  group_by(ID, YEAR) %>% #group by study, year
  summarize(EFFORT.YEAR = n_distinct(PLOT)) %>% #count number of distinct plots 
  left_join(filtered.collated.pairs, ., by = c('ID', 'YEAR')) %>% 
  mutate(ST.ABUN = sum.allrawdata.ABUNDANCE/EFFORT.YEAR) %>% 
  mutate(ST.BIO = sum.allrawdata.BIOMASS/EFFORT.YEAR)

```




### 2.2. Calculate metrics with srtandarized abundance and biomass

Calculate mean, median, min, max, sd, and CoV with standarized abundance and biomass:
- Only genera and study ID kept (not Lat/Long), which standardizes Lat/Long within each study
- Aggregated species to the genus level
- tested to make sure quantile calculation is correct

```{r}

collated.pairs_standardized_summary <- collated.pairs_standardized %>% #read in filtered collated pairs
  group_by(ID,YEAR,GENUS,TAXA, ORGANISMS, CLIMATE, REALM) %>% #group by study, location, genus
  summarize(mean_abun_st=mean(ST.ABUN,na.rm=T), #mean abundance
            median_abun_st=median(ST.ABUN,na.rm=T), #median abundance
            min_abun_st=min(ST.ABUN), # min abundance
            avg_abun_min_5perc = mean(ST.ABUN[ST.ABUN <= quantile(ST.ABUN, .05, na.rm = T)], 
                                     na.rm=T), # calculate specified quantile (in this case lowest 5% of data), and then calc mean of these values
            max_abun_st=max(ST.ABUN), # max abundance
            avg_abun_max_5perc = mean(ST.ABUN[ST.ABUN <= quantile(ST.ABUN, .95, na.rm = T)], 
                                     na.rm=T), # mean of top 5% values, abundance
            sd_abun_st=sd(ST.ABUN,na.rm=T), #sd abundance (when more than one species per genera)
            CoV_abun_st = sd_abun_st/mean_abun_st, # coef of variation
            mean_bio_st=mean(ST.BIO,na.rm=T), #mean biomass
            median_bio_st=median(ST.BIO,na.rm=T), #median biomass
            min_bio_st=min(ST.BIO), #min biomass
            avg_bio_min_5perc = mean(ST.BIO[ST.BIO <= quantile(ST.BIO, .05, na.rm = T)], 
                                         na.rm=T), # mean lowest 5% biomass
            max_bio_st=max(ST.BIO), #max biomass
            avg_bio_max_5perc = mean(ST.ABUN[ST.BIO <= quantile(ST.BIO, .95, na.rm = T)], 
                                          na.rm=T), # mean of top 5% values, biomass
            sd_bio_st=sd(ST.BIO,na.rm=T), #sd biomass
            CoV_bio_st = sd_bio_st/mean_bio_st) #coef of variation


#save(collated.pairs_standardized_summary, file = "data/prep_biotime/collated.pairs_standardized_summary_GLL.Rdata")

```









#3. Find overlapping consecutive years for all pairs

```{r}
bio.pairs <- read_csv("data/prep_biotime/bio_pairs_10km.csv")
```



#### Run for BETWEEN studies

```{r}

pb<-set.prog.bar(nrow(bio.pairs)) #sets progress bar
between.studies.overlap=data.frame(matrix(ncol=5,nrow=0, dimnames=list(NULL, c("Gn1", "Gn2", "Max.Overlap","Type","PAIR.ID")))) #makes an empty dataframe
for (h in 1:nrow(bio.pairs)){
  pb$tick()
  data.pair<-calc.overlap.between(h,collated.pairs_standardized_summary)
  between.studies.overlap<-rbind(between.studies.overlap,data.pair)
  
} #populates dataframe with pairs

```



#### Run for WITHIN studies

```{r}

pb<-set.prog.bar(length(unique(c(bio.pairs$ID.1,bio.pairs$ID.2)))) #sets progress bar
within.studies.overlap=data.frame(matrix(ncol=5,nrow=0, dimnames=list(NULL, c("Gn1", "Gn2", "Max.Overlap","Type","PAIR.ID")))) #makes an empty dataframe
for (h in 1:length(unique(c(bio.pairs$ID.1,bio.pairs$ID.2)))){
  #pb$tick()
  id=unique(c(bio.pairs$ID.1,bio.pairs$ID.2))[h]
  data.pair<-calc.overlap.within(id,collated.pairs_standardized_summary)
  within.studies.overlap<-rbind(within.studies.overlap,data.pair)
  
} #populates dataframe with pairs

```



#### Combine Between and Within:

```{r}

overlap.all.pairs<-rbind(between.studies.overlap,within.studies.overlap)

#saveRDS(between.studies.overlap,"data/data_processing/between.studies.overlap.RDS")

```


Load previously saved version

```{r}

between.studies.overlap <- readRDS("./data/data_processing/between.studies.overlap.RDS")

head(between.studies.overlap)

```








# 4. Calculate portion change for species pairs

Load data

```{r}

load("../data/data_processing/outdated/cleaned_collated_standardized_MSF.Rdata") #NOTE WE ARE LOADING OUTDATED DATA - CHECK WHAT DATA IS REPLACING IT (Isac?)
between.studies.overlap<-readRDS("../data/data_processing/between.studies.overlap.RDS")
within.studies.overlap<-readRDS("data/data_processing/within.studies.overlap.RDS")

bio.pairs <- read.csv("data/prep_biotime/bio_pairs_10km.csv")
meta.pairs <- read.csv("data/prep_biotime/meta_pairs_10km.csv")

```


Select only pairs that overlap for a minimum of 10 years

```{r}

pairs.keep=between.studies.overlap[which(between.studies.overlap$Max.Overlap>9),] #this is generated in the Overlap script

head(pairs.keep)

```


### 4.1 Get proportion change for every genera pair:

```{r}

results=data.frame(matrix(ncol=14,nrow=0, dimnames=list(NULL, c("Gn1", "Gn2", "Log.prop.change.abun.Gn1","Log.prop.change.abun.Gn2","Log.prop.change.bio.Gn1","Log.prop.change.bio.Gn2","PairID","Type","SERIES.n","SERIES.start","SERIES.end","SERIES.l","YEAR.T","YEAR.T1")))) #makes an empty dataframe


pb<-set.prog.bar(nrow(pairs.keep))

for (i in 1:nrow(pairs.keep)){
  pb$tick()
  wk<-get.log.prop.change(i,collated.pairs_standardized_summary,pairs.keep)
  results<-rbind(results,wk)
  
}


head(results)

```



### 4.2. Add metadata:

```{r}

meta.data<-make.meta(results,collated.pairs_standardized_summary)

```



log.prop.change.with.meta rename it for log.prop.change.with.meta.BETWEEN

The code is the same for both Between and Within.

Run it twice for each and then merge them

```{r}

log.prop.change.with.meta.WITHIN<-left_join(results,meta.data) #join meta data with results df

#add unique genera ID col
log.prop.change.with.meta$UNIQUE.PAIR.ID=paste(log.prop.change.with.meta$Gn1,log.prop.change.with.meta$Gn2,log.prop.change.with.meta$PairID,sep="_")
log.prop.change.with.meta.WITHIN$UNIQUE.PAIR.ID=paste(log.prop.change.with.meta.WITHIN$Gn1,log.prop.change.with.meta.WITHIN$Gn2,log.prop.change.with.meta.WITHIN$PairID,sep="_")

#saveRDS(log.prop.change.with.meta,"data/log.prop.change.with.meta.RDS")
#saveRDS(log.prop.change.with.meta.WITHIN,"data/preprocessing/log.prop.change.with.meta.WITHIN.RDS")


```


```{r}
log.prop.change.with.meta<-readRDS("data/log.prop.change.with.meta.RDS")
```




Add better taxa cols

```{r}

taxa1<-tax_name(query = unique(log.prop.change.with.meta$Gn1), get = "class", db = "itis")
taxa2<-tax_name(query = unique(log.prop.change.with.meta$Gn2)[-which(unique(log.prop.change.with.meta$Gn2)%in%unique(log.prop.change.with.meta$Gn1))], get = "class", db = "itis")
taxa<-rbind(taxa1,taxa2)
#add new taxa col
taxa<-readRDS("data/class.assignment.RDS")
taxa$Gn1=taxa$query
taxa$Gn2=taxa$query
names(taxa)[which(names(taxa)=="class")]<-"RESOLVED.TAXA1"
taxa$RESOLVED.TAXA2=taxa$RESOLVED.TAXA1

log.prop.change.with.meta.taxa<-left_join(log.prop.change.with.meta,taxa[,c("Gn1","RESOLVED.TAXA1")])
log.prop.change.with.meta.taxa<-left_join(log.prop.change.with.meta.taxa,taxa[,c("Gn2","RESOLVED.TAXA2")])

log.prop.change.with.meta.WITHIN<-left_join(log.prop.change.with.meta.WITHIN,taxa[,c("Gn1","RESOLVED.TAXA1")])
log.prop.change.with.meta.WITHIN<-left_join(log.prop.change.with.meta.WITHIN,taxa[,c("Gn2","RESOLVED.TAXA2")])


#saveRDS(log.prop.change.with.meta.taxa,"data/preprocessing/log.prop.change.with.meta.w.taxa.RDS")
#saveRDS(log.prop.change.with.meta.WITHIN,"data/preprocessing/log.prop.change.with.meta.WITHIN.w.taxa.RDS")

```


```{r}

table(log.prop.change.with.meta.WITHIN$RESOLVED.TAXA1)
table(log.prop.change.with.meta.WITHIN$RESOLVED.TAXA2)

```


Combine both and split for biomass and abundance

```{r}

between<-readRDS("data/preprocessing/log.prop.change.with.meta.BETWEEN.w.taxa.RDS")
witin<-readRDS("data/preprocessing/log.prop.change.with.meta.WITHIN.w.taxa.RDS")

```

```{r}

all=rbind(between,witin)
remove(witin,between)

```



split again for biomass and abundance

```{r}

abun=all[which(!is.na(all$Log.prop.change.abun.Gn1) & !is.na(all$Log.prop.change.abun.Gn2)),-which(names(all)%in%c("Log.prop.change.bio.Gn1","Log.prop.change.bio.Gn2"))] #keep only rows with non NA abundance values and get rid of bio cols
bio=all[which(!is.na(all$Log.prop.change.bio.Gn1) & !is.na(all$Log.prop.change.bio.Gn2)),-which(names(all)%in%c("Log.prop.change.abun.Gn1","Log.prop.change.abun.Gn2"))] #keep only rows with non NA biomass values and get rid of abun cols

cross.abun.bio=all[which(!is.na(all$Log.prop.change.abun.Gn1) & !is.na(all$Log.prop.change.bio.Gn2)),-which(names(all)%in%c("Log.prop.change.bio.Gn1","Log.prop.change.abun.Gn2"))]
cross.bio.abun=all[which(!is.na(all$Log.prop.change.bio.Gn1) & !is.na(all$Log.prop.change.abun.Gn2)),-which(names(all)%in%c("Log.prop.change.abun.Gn1","Log.prop.change.bio.Gn2"))]

names(abun)[which(names(abun)=="Log.prop.change.abun.Gn1")]<-"Prop.Change.Gn1"
names(abun)[which(names(abun)=="Log.prop.change.abun.Gn2")]<-"Prop.Change.Gn2"

names(bio)[which(names(bio)=="Log.prop.change.bio.Gn1")]<-"Prop.Change.Gn1"
names(bio)[which(names(bio)=="Log.prop.change.bio.Gn2")]<-"Prop.Change.Gn2"

names(cross.abun.bio)[which(names(cross.abun.bio)=="Log.prop.change.abun.Gn1")]<-"Prop.Change.Gn1"
names(cross.abun.bio)[which(names(cross.abun.bio)=="Log.prop.change.bio.Gn2")]<-"Prop.Change.Gn2"

names(cross.bio.abun)[which(names(cross.bio.abun)=="Log.prop.change.bio.Gn1")]<-"Prop.Change.Gn1"
names(cross.bio.abun)[which(names(cross.bio.abun)=="Log.prop.change.abun.Gn2")]<-"Prop.Change.Gn2"

abun$Metric="ABUNDANCE"
bio$Metric="BIOMASS"
cross.abun.bio$Metric="CROSS"
cross.bio.abun$Metric="CROSS"

full.data<-rbind(abun,bio,cross.abun.bio,cross.bio.abun)



#saveRDS(full.data,"data/data_processing/log.prop.change.full.data.RDS")

```


Add distance col

```{r}

full.data=readRDS("data/data_processing/log.prop.change.full.data.RDS")
bio.pairs=read.csv("data/prep_biotime/bio_pairs_10km.csv")

bio.pairs$PairID=paste(bio.pairs$ID.1,bio.pairs$ID.2,sep="_")
full.data<-left_join(full.data,bio.pairs[,c(1,22)])
full.data$dist[which(full.data$ID1==full.data$ID2)]<-0

#saveRDS(full.data,"data/data_processing/log.prop.change.full.data.RDS")

```






# 5. Assign interaction types between pairs



```{r}
log_change <- log.prop.change.with.meta
```


! [[[ DOUBT: shouldn't it read the new "log.prop.change.full.data" instead of "log.prop.change.with.meta"? -- clarify with **Emily** ]]]




list of bio pairs with at least 10 overlapping years:

```{r}

genus_pairs <- apply(log_change[, c("Gn1", "Gn2")], 1, function(x) paste(sort(x), collapse = "-"))
unique_pairs <- unique(genus_pairs)
split_pairs <- data.frame(Gn1 = sapply(strsplit(unique_pairs, "-"), "[", 1),
                          Gn2 = sapply(strsplit(unique_pairs, "-"), "[", 2))
```


Short sample dataset:

```{r}

# Split into 5 groups of roughly equal size
groups <- cut(seq_along(1:12623), breaks = 5, labels = FALSE)

# Split data into separate data frames
df_split <- split(split_pairs, groups)
names(df_split) <- paste0("split_", names(df_split))


list2env(df_split, envir = .GlobalEnv)

```


### BETWEEN studies

```{r}


pair_interactions_1.1 <- make_split(split = split_1)

pair_interactions_2 <- make_split(split = split_2)

pair_interactions_2 <- make_split(split = split_3)

pair_interactions_2 <- make_split(split = split_4)

pair_interactions_2 <- make_split(split = split_5)


#write.csv(pair_interactions_1, "data/data_processing/pair_interactions_1.csv")
#write.csv(pair_interactions_2, "data/data_processing/pair_interactions_2.csv")
#write.csv(pair_interactions_3, "data/data_processing/pair_interactions_3.csv")
#write.csv(pair_interactions_4, "data/data_processing/pair_interactions_4.csv")
#write.csv(pair_interactions_5, "data/data_processing/pair_interactions_5.csv")


```


```{r}

all_interactions <- rbind(pair_interactions_1, pair_interactions_2, pair_interactions_3, 
                          pair_interactions_4, pair_interactions_5)
colnames(all_interactions) <- c("Gn1", "Gn2", "interaction")

#write.csv(all_interactions, "data/data_processing/genus_interaction_list.csv")

```


Read in all_interactions_2:


```{r}


all_interactions <- read.csv("data/data_processing/genus_interaction_list.csv")

```

```{r}

parasite <- all_interactions %>%
  filter(interaction=="pathogenOf")

```




Remove interaction pairs that are mistakes (pathogen, host, hemiparasite - see original all interactions above):

```{r}

mistake_interactions <- c("hasHost", "pathogenOf", "hemiparasiteOf", 'hostOf', 'pathogenOf', 
                          'hasVector', 'parasiteOf')
all_interactions_2<- all_interactions %>%
  filter(!interaction %in% mistake_interactions)
# write.csv(all_interactions_2, "data/data_processing/genus_interaction_list.csv")

summary_interactions <- all_interactions_2 %>%
  group_by(Gn1, Gn2) %>%
  summarize(n=n()) %>%
  arrange(desc(n))

#unique rows summary_interactions: 
unique_interactions_1 <-summary_interactions %>%
  select(Gn1, Gn2) %>%
  distinct()

```



Next steps: start collapsing interaction types into larger groups, 
-predator-prey (including herbivory)

-eatenBy, eats, preysOn, preyedUponBy
-mutualism
-mutualistOf
-dispersal 
-hasDispersalVector, dispersalVectorOf
-visits
-flowersVisitedBy, #visits, #visitsFlowersOf, #visitedBy
-uncategorized_interaction
-interactsWith


```{r}

cased_interactions<- all_interactions_2%>%
  mutate(
    interaction_type = case_when(
      interaction %in%  c("preysOn", "preyedUponBy", "eatenBy", "eats") ~ "predator_prey",
      interaction %in% c("mutualistOf") ~ "mutualism",
      interaction %in% c("hasDispersalVector", "dispersalVectorOf") ~ "dispersal",
      interaction %in% c("interactsWith", "flowersVisitedBy", "visits", "visitsFlowersOf", "visitedBy") ~ "uncategorized_interaction",
      TRUE ~ NA_character_
    ),
    .keep = "unused"
  )

```



Remove pairs whose characters don't match anything:

```{r}

cased_interactions <- cased_interactions %>%
  filter(!is.na(interaction_type))
cased_interaction_unique <- cased_interactions %>% distinct()

```


Check how many have multiple interactions still assigned:

```{r}

summary_interactions_cased<- cased_interaction_unique %>%
  group_by(Gn1, Gn2) %>%
  summarize(n=n()) %>%
  arrange() %>%
  left_join(cased_interaction_unique, by=c('Gn1', "Gn2"))

```



Remove rows with "unclassified_interaction" that are associated with any other interaction type for each genus pair

```{r}

#cased_interactions_filtered <- summary_interactions_cased %>%
# filter(!(n>1 & interaction_type =="uncategorized_interaction"))
# cased_interactions_filtered <- summary_interactions_cased %>%
#   filter(!(n>1 & interaction_type =="visits"))

cased_interactions_filtered_2<- cased_interactions %>%
  group_by(Gn1, Gn2) %>%
  mutate(num_interactions=n())

```



To-do: 

-yay-nay interaction
-Positive-negative-neutral
-Integrate with existing dataset of log abundance for modellers and push 
-Remove original n column from finalized dataset of interactions

-Yes/no interaction
-Get pairs 

```{r}

distinct_pairs <- cased_interactions_filtered_2 %>%
  select(Gn1, Gn2) %>%
  distinct()

```

```{r}

#Read in interactions 
log_change <- readRDS("data/data_processing/log.prop.change.with.meta.RDS")
head(log_change)

#Create a new column in df2 called "interaction" and initialize all values to 0

log_change$interaction_present <- 0

```


Loop through each row in df2 and check if it matches any pairs in df1

```{r}

pb<-set.prog.bar(nrow(log_change)) #sets progress bar
for (i in 1:nrow(log_change)) {
  pb$tick()
  if (paste(log_change$Gn1[i], log_change$Gn2[i], sep = "-") %in% 
      paste(distinct_pairs$Gn1, distinct_pairs$Gn2, sep = "-")
      | paste(log_change$Gn2[i], log_change$Gn1[i], sep = "-") %in% 
      paste(distinct_pairs$Gn1, distinct_pairs$Gn2, sep = "-")) {
    log_change$interaction_present[i] <- 1
  }
}

colnames(log_change)[26] <- "interaction_found"

#saveRDS(log_change, "data/data_processing/log.prop.change.interactions.RDS")

```



Do positive negative neutral 

```{r}

log_change_interaction <- readRDS("data/data_processing/log.prop.change.interactions.RDS")

```


```{r}

pos_neg_interactions<- cased_interactions_filtered_2%>%
  mutate(
    interaction_benefit = case_when(
      interaction_type %in%  c("predator_prey") ~ "negative",
      interaction_type %in% c("mutualism", "dispersal") ~ "positive",
      interaction_type %in% c("uncategorized_interaction") ~ "neutral",
      TRUE ~ NA_character_
    ),
    .keep = "unused"
  )

```

