---
title: "Generating data attributes table"
output: html_notebook
---

This script produces the data attributes table, which shows the studies used in the analyses and their metadata.

We will combine information from the processed data used in the analyses (number and identity of genus taxa, and presence/absence of treatment) with the Biotime metadata.



```{r}
library(ggpubr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
```


## Load data

```{r}
# processed data used for the analyses

load(file = "../../output/prep_data/model_data_final.Rdata")

# Biotime metadata

metadata <- read.csv(file = "../../../data/prep_biotime/BioTIMEMetadata_24_06_2021.csv")

head(moddat)

```



Separate genus code into two columns

```{r}
moddat <- moddat %>%
  rowwise() %>%
  mutate(
    G1 = strsplit(as.character(TS_ID), "~")[[1]][1],
    G2 = strsplit(as.character(TS_ID), "~")[[1]][2]
  ) %>%
  ungroup()

head(moddat)


```


Compute two new information columns for the data used in the analyses: the number ("n_genus") and names ("string_taxa") of the genus in a given study:


```{r}

# create empty columns

moddat$string_taxa <- character(nrow(moddat))
moddat$n_genus <- numeric(nrow(moddat))


# fill in columns

for (i in 1:length(unique(moddat$STUDY_ID))) {
  
  study_id <- unique(moddat$STUDY_ID)[i] # select a study ID
  
  # Compute vector of taxa names
  
  
  vec_g1 = unique(moddat[moddat$STUDY_ID == study_id, "G1"])$G1
  vec_g2 = unique(moddat[moddat$STUDY_ID == study_id, "G2"])$G2
  
  
  
  string_list_taxa =  paste(unique(c(vec_g1, vec_g2)), collapse = ", ") 

  
  moddat[which(moddat$STUDY_ID == study_id), "string_taxa"] = string_list_taxa
  
  
   # Compute n taxa
  
  n_genus <- length(unique(c(vec_g1, vec_g2)))
  
  moddat[which(moddat$STUDY_ID == study_id), "n_genus"]  = n_genus
  
}



```



## Create data attributions table

We use the data used for the analyses and the metadata to create the attributions table:


```{r}

# Select studies from the metadata table used for the analyses

metadata_filtered <- metadata[which(metadata$STUDY_ID %in% moddat$STUDY_ID),]


if(length(unique(moddat$STUDY_ID)) ==  length(unique(metadata_filtered$STUDY_ID))){
  
  print("the number of studies from the filtered metadata matches those from the analyses")
  
}else{
  print("problem")
}

```


Add string_taxa, n_genus and tratment_yn_clean to the metadata:


```{r}
# add variables

metadata_filtered$STRING_TAXA <- character(nrow(metadata_filtered))
metadata_filtered$DISTURBANCE <- character(nrow(metadata_filtered))
metadata_filtered$N_GENERA <- numeric(nrow(metadata_filtered))


# add the computed information columns

for (i in 1:nrow(metadata_filtered)) {
  
  # string taxa
  
  study <- metadata_filtered$STUDY_ID[i]

metadata_filtered$STRING_TAXA[i] <- as.character(moddat[which(moddat$STUDY_ID == study)[1], "string_taxa"])

metadata_filtered$DISTURBANCE[i] <- as.character(moddat[which(moddat$STUDY_ID == study)[1], "treatment_yn_clean"])

metadata_filtered$N_GENERA[i] <- as.numeric(moddat[which(moddat$STUDY_ID == study)[1], "n_genus"])

  
}

## add an extra column indicating the number of genera pairs

metadata_filtered$N_GENERA_PAIRS <- as.numeric(metadata_filtered$N_GENERA)*as.numeric(metadata_filtered$N_GENERA)


# Select the data to show in the table

attributions_table <- metadata_filtered[,
                                                        c("STUDY_ID",
                                                          "WEB_LINK",
                                                          "START_YEAR",
                                                          "END_YEAR",
                                                          "CENT_LAT",
                                                          "CENT_LONG",
                                                          "ORGANISMS",
                                                          "NUMBER_OF_SPECIES",
                                                          "N_GENERA",
                                                          "N_GENERA_PAIRS",
                                                          "STRING_TAXA",
                                                          "AREA_SQ_KM",
                                                          "DISTURBANCE",
                                                          "CONTACT_1",
                                                          "CONTACT_2",
                                                          "CONT_1_MAIL",
                                                          "CONT_2_MAIL",
                                                          "LICENSE",
                                                          "DATA_SOURCE")]

attributions_table

# save table

write.csv(attributions_table, "attributions_table2.csv")
```

